NAND formulas lower bounds

### 2021-06-10 nand_lb.tex
Slowdown Lemam

\newcommand{\NN}{\mathbb{N}}%
\newcommand{\Sbb}{\mathbb{S}}%
\newcommand{\Cc}{\mathcal{C}}%
\newcommand{\Gg}{\mathcal{G}}%
\newcommand{\Ss}{\mathcal{S}}%
\newcommand{\Ot}{\tilde{O}}%
\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\bit}{\{0,1\}}%
\DeclareMathOperator*{\poly}{poly}

@@

	In mathematical terms: we associate to each node of the OR/AND tree a bit string of length $\leq n$ 
	in the following way: 
	start with the empty string at the root, append a zero when you move to a left child, and a one when you move to a right child. Leaves have bit strings of length $n$.
	Given a leaf with bitstring $y$, we add below it the formula 
	$((y = x) \wedge f(y)) \vee ((y\neq= x) \wedge y fiddle x)$ TODO.
\end{proof}

@@

\section{To keep in mind for later}
It seems that function similar to $x_0 \vee (x_1 \wedge (x_2 \vee \ldots))$
are ``hard'', i.e. they have NAND formulas of depth at least $n$ (or $\Omega(n)$ ?).

\begin{itemize}
	\item Is this true ?
	\item Can we formalize it ?
	\item Can we use it to find another hard function like Andreev's ?
\end{itemize}

>>>>

\newcommand{\NN}{\mathbb{N}}%
\newcommand{\Sbb}{\mathbb{S}}%
\newcommand{\Cc}{\mathcal{C}}%
\newcommand{\Gg}{\mathcal{G}}%
\newcommand{\Ss}{\mathcal{S}}%
\newcommand{\Ot}{\tilde{O}}%
\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\bit}{\{0,1\}}%
\DeclareMathOperator*{\poly}{poly}
\DeclareMathOperator*{\polylog}{polylog}

@@

	In mathematical terms: we associate to each node of the OR/AND tree a bit string of length $\leq n$ 
	in the following way: 
	start with the empty string at the root, append a zero when you move to a left child, and a one when you move to a right child. Leaves have bit strings of length $n$.
	Given a leaf with bit string $y$, we add below it the formula 
	$((y = x) \wedge f(y)) \vee ((y\neq= x) \wedge y fiddle x)$ TODO.
\end{proof}

@@

\section{Speedup Lemma}

\begin{lemma}\label{speedup-lemma}
	Let $F$ be a uniform OR/AND formula on $n$ variables of depth $d= c \log n$.
	There is an (co-)nondeterministic algorithm that decides
	whether an input $v$ satisfies $F$ in time $n^{c/2}\polylog(n)$.
\end{lemma}
\begin{proof}
	We first prove the nondeterministic case.
	We can assume w.l.o.g. that $d$ is even, and $F$ has depth $d$ on every path.

	Given an OR/AND formula $F$ of depth $d$ a bit string $x$ of length at most $d$, let $F_{|x}$
	be the sub-formula defined by the subtree starting at the node reached by following the path specified by $x$ from the root or $F$.

	Consider the following recursive algorithm $A$, taking as input an OR/AND formula $F$ of depth $d > 0$, a bit string $x$, and the valuation $v$:
	\begin{itemize}
		\item 
	if $|x| < d$
	nondeterministically guess $b\in \bit$,
	and return $A(F, xb0, v) \wedge A(F, xb1, v)$.
		\item 
	if $|x| = d$, run the the algorithm $P_F$ on $x$, and return the truth value of the corresponding literal in $v$.
	\end{itemize}

	Correctness: we show that there exists a $b\in \bit$ such $A(F, x, v)$ returns true iff and only if
	the formula $F_{|x}(v)$ is true.
	This is follows by definition when $|x| = d$.
	By induction, when $|x| < d$, $A(F, xb0, v) \wedge A(F, xb1, v)$ is the value of the AND/OR sub-formula $F_{|xb}$. As $F_{|x}$ is an OR/AND formula, it is satisfiable iff at least one of $F_{|x0}$ or $F_{|x1}$ is sat, i.e. iff there exists $b\in \bit$ such that $A(F, xb0, v) \wedge A(F, xb1, v)$ is true.
	The value of $F$ is then given by $A(F, \varepsilon, v)$, where $\varepsilon$ is the empty word.

	Complexity: $A$ makes two recursive calls with the length of $x$ increased by $2$, 
	and makes a call to $P_F$ at length $d$.
	Therefore there are $2^{d/2} = n^{c/2}$ calls to $P_F$, and $A$ runs in time $n^{c/2}\polylog(n)$.
	We also see here that increasing the depth of $F$ by a constant only adds a constant factor in the runtime, hence we can assume that $F$ has even depth.

	For the co-nondeterministic case, we proceed similarly, except that we guess universally on the AND gates,
	and check both sides of OR gates.
\end{proof}

\section{Lower bounds}

\section{To keep in mind for later}
It seems that functions of the form $x_0 \vee (x_1 \wedge (x_2 \vee \ldots))$
are ``hard'', i.e. they have NAND formulas of depth at least $n$ (or $\Omega(n)$ ?).

\begin{itemize}
	\item Is this true ?
	\item Can we use it to construct a hard function like Andreev's ?
\end{itemize}
